Binary companions of red giant stars with APOGEE: \\ catalog and companions that survive the common envelope

### 2018-01-23 paper/1-catalog/twoface-catalog.tex
more text

% define macros for text
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\sdssiii}{\project{\acronym{SDSS-III}}}
\newcommand{\thejoker}{\project{The~Joker}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\DR}{\acronym{DR14}}
\newcommand{\RC}{\acronym{RC}}
\newcommand{\RGB}{\acronym{RGB}}

@@

Stars typically have companions.

Main sequence stars in the solar neighborhood more often appear in binary or
multiple star systems rather than as solitary stars
(\citealt{ragavhan,tokovinin,todo}).
This is likely a generic outcome of star formation: turbulent fragmentation in
collapsing protostellar clouds tends to produce multiplets of stars within
length-scales comparable to their spheres of influence (e.g.,
\citealt{Raskutti:2016}).
Binary and multiple star systems are therefore of great interest in
astrophysics: the population of stars and their companions encodes information
about star formation processes (\citealt{todo}), stellar parameters and
evolution (\citealt{todo}), and the dynamics of multi-body systems
(\citealt{todo}).

Even without stellar or sub-stellar companions, single stars likely have
planetary-mass companions (\citealt{todo}).
\todo{Planetary systems are interesting because...yadda yadda}
Differences between the population of stellar companions and that of
planetary-mass companions likely relate directly to the different formation
channels for these objects.

\todo{should we name ``stars with companions'' something like ``stellar
systems''?}
Stars with companions have a large dynamic range of constituent and orbital
characteristics.
For example, for binary main sequence stars, mass-ratios span from XX to 1 and
periods from XX to YY years.
For planetary companions of main sequence stars, \todo{}.
- Some indication of trends (e.g., with metallicity) -> star and
  planet formation processes

Most of what is known about stellar companions comes from main sequence stars.
- Some evolved stars have special roles in exotic systems like cataclysmic
  variables (donor stars)
- But less is known about the population of companions to evolved stars.
- Expect differences in evolved star pop. from main sequence populations:
  - mass loss
  - envelope -> engulfment?
- Current samples limited in dynamical range of chemical properties, location,
  age (local to sun)
- Brighter, but typically larger distances, direct imaging not feasible
- Eclipsing companions good, but low mass companions stars -> hard to find
  (giants are big)
- Need large sample of stars with time-domain spectroscopy

@@

We have solved this problem previously, though with limitations (to be discussed
more below), with \thejoker\ (\citealt{Price-Whelan:2017}).
\thejoker\ is a Monte Carlo rejection sampler that is computationally expensive
but probabilistically righteous:
It delivers independent posterior pdf samples for single-companion binary
orbital parameters, given any number of radial-velocity measurements.
Here we use \thejoker\ to generate posterior pdf samples for \todo{which?} stars
observed by the \apogee\ surveys (see \sectionname~\ref{sec:data};
\citealt{Majewski:2015}).

The \apogee\ surveys primarily target red-giant-branch (\RGB) stars, which are
ideal for the study of single-line binary systems.
For one, because they are so luminous, they are unlikely (in general) to have
equally-bright companions, and their spectra are therefore well-approximated or
fit as single-line objects.
The subset of \RGB\ stars in the ``red clump'' (\RC) are even more powerful as
they are standard candles and have masses that can be estimated using
spectroscopy (using dredged-up elements; \citealt{Martig:2016,Ness:2016}).
With primary-star mass estimates, the binary-orbit fitting will return
$m_2\,\sin i$ estimates for the secondary, and not just estimates of the
mass-function.
Additionally, the \apogee\ pipelines (\citealt{Garcia-Perez:2016}) and also
\thecannon\ (\citealt{Ness:2015}) produce detailed abundance estimates for \RGB\
and \RC\ stars.
If there are causal relationships between chemical abundances and binary
companions---as are expected---these should be measurable here.

By making cuts on this library of posterior pdf samples (described in detail in
\sectionname~\ref{sec:whatever}), we deliver a catalog of binary-star systems
from the \apogee\ survey with no cuts or constraints on data quality or volume.
We additionally highlight the subset of this catalog with \RC\ star primaries,
for which we can compute companion mass and pericentric distance.

@@

\todo{HOGG: fill in this ish}

HOGG: Why and how does \apogee\ rock it?

ASPCAP: \citealt{Garcia-Perez:2016}
SDSSIV: \citealt{Blanton:2017}

Overview of what \apogee\ is, and what data we are using.

How we updated the individual-visit radial velocity measurements.

What we did (if anything) with the missing visits.

>>>>

% define macros for text
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\sdssiv}{\project{\acronym{SDSS-IV}}}
\newcommand{\thejoker}{\project{The~Joker}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\DR}{\acronym{DR14}}
\newcommand{\RC}{\acronym{RC}}
\newcommand{\RGB}{\acronym{RGB}}

@@

Stars typically have companions. Main sequence stars in the solar neighborhood
more often appear in binary or multiple star systems rather than as solitary
stars (\citealt{ragavhan,tokovinin,todo}).
This is likely a generic outcome of star formation: turbulent fragmentation in
collapsing protostellar clouds tends to produce multiplets of stars within
length-scales comparable to their spheres of influence (e.g.,
\citealt{Raskutti:2016}).
Binary and multiple star systems are therefore of great interest in
astrophysics: the population of stars and their companions encodes information
about star formation processes (\citealt{todo}), stellar parameters and
evolution (\citealt{todo}), and the dynamics of multi-body systems
(\citealt{todo}).

% Even without stellar or sub-stellar companions, the vast majority of stars have
% planetary-mass companions (\citealt{todo}).
% \todo{Planetary systems are interesting because...blah}
% Differences between the population of stellar companions and that of
% planetary-mass companions likely relate directly to the different formation
% channels for these objects.

Most of what is known about stellar companions comes from studies of
main-sequence (MS) stars.
MS stars with companions have a large dynamic range of constituent and orbital
characteristics.
For example, binary stars have mass-ratios that span from $\approx 10^-3$
\todo{citation because this is made up} to 1, and have periods from days to
millions of years.
\todo{Describe distributions over these quantities}
\todo{multiplicity in general: abundance, numbers, etc.}
\todo{some other stats from Moe and Di Stefano}

Some evolved stars have special roles in well-studied exotic multiple systems
like dwarf novae or cataclysmic variables \todo{are those examples correct?},
but by-in-large, less is known about the population of companions to evolved
stars.

- Expect differences in evolved star pop. from main sequence populations:
  - mass loss
  - envelope -> engulfment?
- Current samples limited in dynamical range of chemical properties, location,
  age (local to sun)
- Brighter, but typically larger distances, direct imaging not feasible
- Eclipsing companions good, but low mass companions stars -> hard to find
  (giants are big)
- Need large sample of stars with time-domain spectroscopy

% For planetary companions of main sequence stars, \todo{}.
% - Some indication of trends (e.g., with metallicity) -> star and
%   planet formation processes

@@

We have solved this problem previously, though with limitations (to be discussed
more below), with \thejoker\ (\citealt{Price-Whelan:2017}).
\thejoker\ is a Monte Carlo rejection sampler that is computationally expensive
but probabilistically righteous:
It delivers independent posterior pdf samples for single-companion binary
orbital parameters, given any number of radial-velocity measurements.
Here we use \thejoker\ to generate posterior pdf samples for stars observed by
the \apogee\ survey (see \sectionname~\ref{sec:data}; \citealt{Majewski:2015}).

The \apogee\ surveys primarily target red-giant-branch (\RGB) stars, which are
ideal for the study of single-line binary systems.
For one, because they are so luminous, they are unlikely (in general) to have
equally-bright companions, and their spectra are therefore well-approximated or
fit as single-line objects.
When this constraint is not met, \thejoker\ will in general fail, and a model
that fits for a mixture of stellar spectra is more appropriate (e.g.,
\citealt{Kareem}).
The subset of \RGB\ stars in the ``red clump'' (\RC) are even more powerful as
they are standard candles and have masses that can be estimated using
spectroscopy (using dredged-up elements; \citealt{Martig:2016,Ness:2016}).
With primary-star mass estimates, the binary-orbit fitting will return
$m_2\,\sin i$ (minimum mass) estimates for the secondary, and not just estimates
of the so-called ``binary mass function.''
Additionally, the \apogee\ pipelines (\citealt{Garcia-Perez:2016}) and also
\thecannon\ (\citealt{Ness:2015}) produce detailed abundance estimates for \RGB\
and \RC\ stars.
If there are causal relationships between chemical abundances and binary
companions---as are expected---these should be measurable.

By making cuts on this library of posterior pdf samples (described in detail in
\sectionname~\ref{sec:whatever}), we deliver a catalog of binary-star systems
from the \apogee\ survey \todo{with ... (tease some statistics)}.
We additionally highlight the subset of this catalog with \RC\ star primaries,
for which we can compute companion mass and pericentric distance.

@@

All data used in this work comes from the publicly-available data release 14
(\DR) of the \apogee\ survey (\citealt{apogeedr14}), a component of the Sloan
Digital Sky Survey IV (\sdssiv; \citealt{sdss,Blanton:2017}).
\apogee\ is an infrared spectroscopic survey designed to map \RGB\ stars across
much of the Milky Way.
Targets are selected with simple color and brightness cuts, but the survey uses
fiber-plugged plates with a maximum of \todo{100?} fibers and a $\approx
1.5~{\rm deg}^2$ field of view, leading to ``pencil-beam''-like sampling of the
stellar distribution.
Data taken as part of the \apogee\ survey is reduced with the \todo{APOGEE
spectral parameter...} (\acronym{ASPCAP}; \citealt{Garcia-Perez:2016}), which
measures stellar parameters, chemical abundances, and radial velocities for each
target.
\todo{talk about visits, how the velocities measured...}

From the \apogee\ \DR\ data products (the \texttt{allStar} and \texttt{allVisit}
files), we select all stars with $\geq 3$ visits that each pass a set of quality
cuts.
For each visit, we require that the following bits are not set in the
\texttt{STARFLAGS} bitmask: \texttt{TODO}.
For each star, we require that the following bits are not set in the
\texttt{ASPCAPFLAGS} bitmask: \texttt{TODO}.
After these cuts, and the requirement of $\geq 3$ visits for a given star, we
are left with \todo{XX} visits for \todo{XX} unique stars (unique
\texttt{APOGEE\_ID}).
\todo{add numpy code for exact filter?}
